---
title: "Associations and trends exercises"
author: "Jon Chernus"
output:
  html_document:
    code_folding: show
---

# Setup

```{r}
# Load packages
library("tidyverse")
library("ggforce")
library("knitr")
library("lubridate")
library("plotly")
library("zoo")
library("spls")
library("GGally")
library("corrplot")
library("plotly")
library("rgl")
library("scatterplot3d")
library("knitr")
library("hexbin")
```

```{r}
# Read in data
cv <- read.csv("https://jmc108.github.io/hugen2073-textbook/exercises/6_associations_and_trends/cv.csv")
lrr <- read.csv("https://jmc108.github.io/hugen2073-textbook/exercises/6_associations_and_trends/lrr.csv")
sc <- read.csv("https://jmc108.github.io/hugen2073-textbook/exercises/6_associations_and_trends/sc.csv")
pcs <- read.csv("https://jmc108.github.io/hugen2073-textbook/exercises/6_associations_and_trends/genetic_data_train.csv")
util <- read_csv("https://jmc108.github.io/hugen2073-textbook/exercises/6_associations_and_trends/utility.csv")
```

# Exercises

## Line graph

The chunk below simulates genetic drift in two populations of different sizes (`Ne`=200 `Ne`=2000).

We start with an initial allele frequency of `p0`=0.3 and follow `p` for 20 generations (`generation`).

Each population is simulated 12 times, as reflected in the grouping variable `pop`. So, for example, `Ne200_7` in the `pop` column means "replicate number 7 of the 200-person population."

Inspect the data to make sure you understand the variables.


```{r}
set.seed(2073)

# ---- Wright–Fisher drift simulator ----
simulate_drift <- function(p0, Ne, gens, n_pops, label_prefix = "pop") {
  # p0: starting allele frequency
  # Ne: effective population size (diploid)
  # gens: number of generations
  # n_pops: number of replicate populations
  
  map_dfr(seq_len(n_pops), function(i) {
    p <- numeric(gens + 1)
    p[1] <- p0
    for (g in 2:(gens + 1)) {
      # next generation allele count ~ Binomial(2Ne, p_prev)
      k <- rbinom(1, size = 2 * Ne, prob = p[g - 1])
      p[g] <- k / (2 * Ne)
    }
    tibble(
      pop = paste0(label_prefix, i),
      generation = 0:gens,
      p = p
    )
  })
}

# ---- realistic-ish parameters ----
p0   <- 0.30   # common-ish starting allele frequency
gens <- 60
n_pops <- 12

# Two scenarios: small vs large Ne to show volatility differences
d_small <- simulate_drift(p0 = p0, Ne = 200,  gens = gens, n_pops = n_pops, label_prefix = "Ne200_")
d_large <- simulate_drift(p0 = p0, Ne = 2000, gens = gens, n_pops = n_pops, label_prefix = "Ne2000_")

d <- bind_rows(
  d_small %>% mutate(Ne = 200),
  d_large %>% mutate(Ne = 2000)
)

kable(head(d))
```

Now make a line graph of the allele frequency over time:

- facet by population size (`Ne`), in such a way that one plot is on top of the other (rather than side by side)

- draw a line for each replicate, but make them slightly translucent with `alpha` (color them all the same; what aesthetic should you use instead of `color`?)

- include `+ theme_classic()` to remove the plot background

- but add a horizontal dotted line to show the initial allele frequency of `p0`=0.3 (what geom do you need? Include `linetype="dashed"`)

```{r}
# Your code here

```

## Bonus: vibe-coded interactive line graph

Your instructor thought his electricity bill looked high, so he downloaded a year's data from the utility company.

This code produces an interactive bar graph of electricity usage every hour, colored by hourly electricity cost.

No need to change anything here - just play with the interactive plot for a few moments and check out how surprisingly simple the plotting code is.

```{r}
d2 <- util %>%
  mutate(
    datetime = as.POSIXct(as.Date(DATE, format = "%m/%d/%y"), tz = "America/New_York") +
      as.numeric(`START TIME`),
    cost = readr::parse_number(COST)
  ) %>%
  arrange(datetime)

p <- plot_ly(
  data = d2,
  x = ~datetime,
  y = ~`USAGE (kWh)`,
  type = "bar",
  customdata = ~cost,
  hovertemplate = paste(
    "%{x}<br>",
    "Usage: %{y:.2f} kWh<br>",
    "Cost: $%{customdata:.2f}",
    "<extra></extra>"
  ),
  marker = list(
    color = ~cost,                 # <- continuous color mapping
    showscale = TRUE,              # <- colorbar legend
    colorbar = list(title = "Cost ($)")
  )
) %>%
  layout(
    title = "Hourly Usage (kWh), Colored by Cost ($)",
    xaxis = list(
      title = "Time",
      type = "date",
      rangeslider = list(visible = TRUE)
    ),
    yaxis = list(title = "Usage (kWh)")
  )

p


```

## Smoothing

The dataset `lrr` (read in above) contains LRR values on chromosome 22 for a single scan.

Make a scatterplot of the data. Then use the function `rollmean` from the `zoo` package to calculate moving average, and plot it as a connected line. Include the option `fill=NA`. Vary the `k` parameter to see if you can find one that does a good job of pinpointing the deletion. (Change `xlim()` if that helps.)

```{r}

```

Try the same thing with a LOESS curve (`geom_smooth(method="loess")`) instead of `rollmean`. Turn off the error bar, and vary the `span` parameter.

```{r}
# Your code here
```


## Overplotting

Now, let's try making a scatterplot of gene expression in two samples in the `sc.csv` dataset (read in above). Pick two samples and make a scatterplot of the expression of all the genes. Apply the `log2(x+1)` transformation first. (Don't use `sample1` and `sample2`, as they're so similar that they make for a boring plot.)

How many points are in the plot? How many points do you think you can see in the plot?

In your first plot, try using the `alpha` parameter to control overplotting.

```{r}
# Your code here

```

Now try a few more ways of addressing the overplotting problem.

First, try out `geom_bin2d` and `geom_hex`

- Try filtering out genes that are unexpressed in either sample

- Also try adjusting bin parameters

```{r}
# Your code here
```

Now try two versions of a 2D density plot: `geom_density_2d` and `geom_density_2d_filled`. Notice the default plot is not helpful without filtering lowly-expressed genes. Try putting the two geoms in a single plot and overlaying a scatterplot.

Note there are some more variations [here](https://ggplot2.tidyverse.org/reference/geom_density_2d.html).

```{r}
# Your code here
```

## Connected scatterplots

If we have two variables that change with time, we could plot a line graph of each. Here, gene-expression measurements from synchronized cells are summarized by two principal components. 

```{r}
# No need to change this code

data(yeast)
# genes x timepoints
dim(yeast$y)
# transpose so rows = timepoints
pca <- prcomp(t(yeast$y), scale. = TRUE)
pca_df <- data.frame(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  idx = seq_len(nrow(pca$x))  # implicit time index
)
pca_df %>% 
  ggplot() +
  geom_point(aes(x=idx, y=PC1)) + geom_line(aes(x=idx, y=PC1), color="red") +
  geom_point(aes(x=idx, y=PC2)) + geom_line(aes(x=idx, y=PC2), color="blue") +
  ylab("PC1 (red) and PC2 (blue)")+
  ggtitle("Two PCs over time")

```


Instead, we could look at how `PC1` and `PC2` relate to each other directly. Make a connected scatterplot to show this.

- Try `geom_line` - why doesn't it work?

- Use `geom_path` instead.

- Try including `arrow=arrow()`.

```{r}
# Your code here

```


## Many variables

We can use the `ggpairs` function from `GGally` to make pairwise plots of many variables.

Try using `ggpairs` to explore expression in 5 samples in the `sc` dataset. `log2(x_1)`-transform the variables and filter out non-expressed genes first.

```{r}
# Your code here
```

## Correlations and heatmaps

Now try using the `corrplot` function from the `corrplot` package.

- Try using the same code you used in the chunk above to filter and process the data

- This time, use all the samples

- Pipe it into the `cor` function to calculate correlations

- Pipe the correlation matrix into `corrplot`

- Experiment with different values of the `method` argument (or use the `corrplot.mixed` function and separately specify `lower` and `upper`; note the default plotting parameters are not wonderful)

```{r}
# Your code here
```


## 3D scatterplot

There are several ways to make 3D scatterplots in R. These include 

- `plotly::plot_ly(type="scatter3d", mode="markers")`

- `rgl::plot3d`

- `scatterplot3d:scatterplot3d` (only a 2D projection)

Try using at least one of the above to make a scatterplot of the first 3 PCs of ancestry in the dataset `pcs`. Color the points by `Ancestry`. You will probably need to check the documentation of the plotting function you use (hint: look for examples at the bottom of the help page).

```{r}
# Your code here
```


## Parallel set/alluvial plot

Let's make a parallel set-style plot to show how the categorical variables `Type` and `clinsig` are related in the `cv` dataset.

There are several packages for these kinds of plots, and they all seem to be a little quirky. We will try `geom_parallel_sets` from the `ggforce` package.

### Formatting

First we need to use some helper functions to get the data in shape for `geom_parallel_set`.

Currently, the data look like this:

```{r}
knitr(head(cv))
```

We need to aggregate the cross-category counts and convert that into parallel sets format.

Copy this code and run it in the chunk below:

```
# Aggregate counts
d_counts <- cv %>%
  count(Type, clinsig, name = "n")

# Convert to parallel-sets format
d_ps <- gather_set_data(d_counts, c("Type", "clinsig"))

head(d_counts)
head(d_ps)
```

```{r}
# Paste the above code here and run it
```

### Plotting

Now we're ready to plot

- Use `ggplot`

  - with `d_ps` as the input data frame
  
  - and `aes(x, id = id, split = y, value = n)`

- Also include these three geoms

  - `geom_parallel_sets`
  
  - `geom_parallel_sets_axes`
  
  - `geom_parallel_sets_labels`

- To improve the plot:

  - adjust `alpha` in one of the layers (which?)
  
  - adjust `size` in one of the layers (which?)
  
  - include `axis.width = 0.25, color = "grey30", fill = "grey95"` in one of the layers (which?)
  
  - try coloring the ribbons by either `Type` or `clinsig` (which layer?)

```{r}
# Your code here
```

