# Setup

Log on to the computing cluster and launch an RStudio Server session.

We will try to use the package `susieR` to reproduce Figure 2 from this paper:

[Gjoka, Aida, and Heather J Cordell. “Fine-Mapping the Results From Genome-Wide Association Studies of Primary Biliary Cholangitis Using SuSiE and h2-D2.” Genetic epidemiology vol. 49,1 (2025): e22592. doi:10.1002/gepi.22592](https://pmc.ncbi.nlm.nih.gov/articles/PMC11656035/).

Summary stats from the GWAS along with LD matrices are available at [https://www.staff.ncl.ac.uk/heather.cordell/GJOKA_SUMSTATS.zip](https://www.staff.ncl.ac.uk/heather.cordell/GJOKA_SUMSTATS.zip). These have already been downloaded for you in `ix1/hugen2073-2026s/susie/GJOKA_SUMSTATS/`.

The locus in Figure 2 is locus number 46, which corresponds to the files `sumstats_46.assoc.logistic` and `covmat_46.ld`.

# Read in the data

```{r}
# install.packages("susieR")
library(data.table)
library(susieR)
library(ggplot2)

# Load your data
# SNP order in sumstats MUST match rows/cols of R - We'll assume this is true
sumstats <- fread("/ix1/hugen2073-2026s/susie/GJOKA_SUMSTATS/sumstats_46.assoc.logistic")
R <- as.matrix(fread("/ix1/hugen2073-2026s/susie/GJOKA_SUMSTATS/covmat_46.ld"))

# Notice the dimensions match
dim(sumstats)
dim(R)

# Inspect
# Looks like plink results
head(sumstats)
R[1:5,1:5]
```

# Run `susieR`

```{r}
# Compute z-scores
z <- sumstats$BETA/sumstats$SE

# Run SuSiE-RSS
fit <- susie_rss(
  z = z,
  R = R,
  n = sumstats$NMISS[1],  # Sample size
  L = 10               # max number of effects
)

# Look at results
# Plot PIPs
summary(fit)
fit$sets
pip <- fit$pip
pip
```

# Plot

```{r}
# Really basic plot
plot(pip)
plot(sumstats$BP, -log10(sumstats$P), col="grey")
points(sumstats$BP, -log10(sumstats$P),
       col=ifelse(pip > 0.1, "red", "grey"))



# Plot the two CSs in different colors
# color CS1
# color CS2
cs <- fit$sets$cs
cols <- rep("grey", nrow(sumstats))
cols[cs[[1]]] <- "red"
cols[cs[[2]]] <- "blue"
plot(sumstats$BP, -log10(sumstats$P), col=cols, pch=16)



# Nicer plot in ggplot
# This plot looks more like Figure 2
cs <- fit$sets$cs
# Assign each SNP to a CS (or "none")
cs_id <- rep("none", nrow(sumstats))
for (i in seq_along(cs)) {
  cs_id[cs[[i]]] <- paste0("CS", i)
}
df <- data.frame(
  BP = sumstats$BP,
  logp = -log10(sumstats$P),
  pip = pip,
  cs = factor(cs_id)
)
ggplot(df, aes(x = BP, y = logp)) +
  geom_point(aes(color = cs, size = pip, alpha = pip)) +
  scale_color_manual(values = c(
    "none" = "grey80",
    "CS1" = "red",
    "CS2" = "blue"
  )) +
  scale_size(range = c(1, 5)) +
  scale_alpha(range = c(0.3, 1)) +
  labs(
    x = "Position (BP)",
    y = expression(-log[10](P)),
    color = "Credible Set",
    size = "PIP",
    alpha = "PIP"
  ) +
  theme_minimal()

```