# Exercises

## Exercise 1

The SNP rs62229686 was the lead variant at a particular locus in a multivariate GWAS of plasma lipid levels. The authors claim it is a missense variant in the AGPAT3 gene with CADD \< 20 (<https://www.nature.com/articles/s41467-023-42532-8>). Try to verify these claims.

(You should notice there's a problem when trying to verify the CADD score through querying with R/SQL. What could you try instead?) You may also find it easier (necessary?) to use two queries for verifying the functional consequence and the gene.

```{r}
# Your code here


```


# Solutions

## Exercise 1

We can pretty easily verify it being a missense variant in AGPAT3.

```{r}
# Get the position and functional consequence
q <- "
SELECT *
FROM hg38.snp151
WHERE name = 'rs62229686'
;"

r <- query(q)
r
r$func

chr_temp <- r$chrom
start_temp <- r$chromStart
end_temp <- r$chromEnd

# Get the gene
q <- paste0("
SELECT *
FROM hg38.ncbiRefSeq
WHERE chrom = '",chr_temp,"' AND
txStart < ",start_temp," AND
txEnd > ",end_temp,"
;"
)
r <- query(q)
table(r$name2)

```

For the CADD part, you can install bigWigToWig and run then it like this. Note that the CADD scores are in 4 separate files, one for each possible allele (A, C, G, T).

```{r}
# You may need to give yourself permission to run the program
# Open your terminal and run this command:
# chmod -v a+x ~/Downloads/bigWigToWig

# This create a bigWigToWig command for us to run
for (allele in c("a","g","t","c")) {

  command_temp <- paste0(
  "~/Downloads/bigWigToWig ",
  "-chrom=",chr_temp," ",
  "-start=",start_temp," ",
  "-end=",end_temp," ",
  "http://hgdownload.soe.ucsc.edu/gbdb/hg38/cadd1.7/",allele,".bw stdout"
  )

  # This causes R to run the command
  system(command_temp)

}
```
